### 2021-10-28 documents/sections/extensions.tex
went through 3.1-3.4

\section{The innovative \gl{1}\gl{s_a_} markers}
\label{sec:extensions}
As stated in \cref{sec:extensions_intro}, the person marker extensions which did not spread through the entire lexicon all have in common that they feature innovative \gl{1}\gl{s_a_} markers.
There are six attested such innovations, three of which can be reconstructed to intermediate proto-languages.
The new \gl{1}\gl{s_a_} prefix is formally identical to the \gl{1+2} marker (\PC \rc{k-}) in three cases, to the \gl{1}\gl{s_p_} marker (\PC \rc{u(j)-}) in two cases, and to the \gl{1}>\gl{3} marker (\PC \rc{t-}) in one case.
\cref{sec:pekodian} investigates the innovation of \rc{k-} in \PPek, reflected in the three daughter languages \arara, \ikpeng, and \bakairi.
\cref{sec:waiwaian} takes a closer look at the extension of \rc{k-} in \PWai, which was briefly shown in \cref{sec:extensions_intro}.
\cref{sec:taranoan} concerns the extension of \rc{t-} in \PTir (or \PTar), reflected in modern \trio and \akuriyo.
\crefrange{sec:akuriyo}{sec:yukpa} look at innovative first person markers which are only attested in single modern languages:
\obj{k-} in \akuriyo, and \obj{j-} in \carijo and \yukpa.

\subsection{\PPek \rc{k-}}
\label{sec:pekodian}
The Pekodian branch was suggested by \textcite{meira2005southern}, as the result of fieldwork on \bakairi by Meira and the availability of more material on \ikpeng. % \parencites{pacheco1997relativizacao}{ikpengpacheco1997}{campetela1997analise}{pacheco1998nominalizacao}{ikpengpacheco2001}{campetela2002aspectos}{pacheco2002aspectos}{pacheco2003intransitivos}.
It consists of closely related \arara and \ikpeng, with \bakairi as a more distant member.
\textcite{meira2005southern} focused on phonological and lexical properties, so no reconstructive work has been done on \PPek morphosyntax.
All three Pekodian languages have a regular \gl{1}\gl{s_a_} marker \obj{k-}, as evidenced by the paradigms in \cref{tab:pekreg}.
Its presence can thus be reconstructed to \PPek.

@@

\begin{multicols}{2}
\ex<arairr> \arara \parencite[153]{alves2017arara} \\
\begin{tabular}[t]{@{}ll@{}}
\obj{wɨ-genɨ} & \qu{I said}\\
\obj{w-it͡ʃinɨ} & \qu{I was, lied down}\\
\obj{w-ebɨnɨ} & \qu{I came}\\
\obj{w-ibɨnɨ} & \qu{I bathed}\\
\obj{w-iptoŋrɨ} & \qu{I went down}\\
\obj{w-ɨdolɨ} & \qu{I went}\\
\end{tabular}
\xe
%
\ex<araap> \arara \parencite[200]{alves2017arara}\\
\begin{tabular}[t]{@{}ll@{}}
\gl{1} & \obj{w-aptam}\\
\gl{2} & \obj{m-od-aptam}\\
\gl{1+2} & \obj{kud-aptam}\\
\gl{3} & ∅\obj{-aptam}\\
\end{tabular}
\xe
\end{multicols}

In his brief discussion of \bakairi verbal person marking, \textcite{meira2003bakairi} reports the existence of two subclasses of \gl{s_a_} verbs, one taking first person \obj{w-}, and one \obj{k-}.
The verb used to illustrate the first group is \obj{i} \qu{to bathe} \exref{bak-33}, contrasting with regular \obj{əku} \qu{to go up} in \exref{pekreg} above.
\qu{to bathe} is also found in the \obj{w-}list for \arara.

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
While \textcite[4]{meira2003bakairi} does list \obj{ge} \qu{to say}, \obj{tə} \qu{to go}, and \obj{əe(wɨ)} \qu{to come} as examples of \gl{s_a_} verbs, he does not indicate whether they belong to the class of \gl{s_a_}-1 verbs, with first person \obj{k-}, or the \gl{s_a_}-2 verbs, with \obj{w-}.\footnote{It should be noted at this point that \textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}. However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}. Given the lack of data for other verbs, we will not further discuss this potential additional pattern. If the characterization by \citeauthor{meira2003bakairi} is accurate, then the pattern is fully parallel to the distribution of the first person prefixes.}
Luckily, while \textcite{von1892bakairi} did not accurately record all phonemic distinctions in \bakairi \parencite{meira2005bakairi}, he does provide inflected forms of cognates to the \arara verbs in \exref{arairr}.
We present them according to our understanding of the phonemic analysis of \bakairi by \textcites{wheatley1969bakairi}{meira2003bakairi}{meira2005bakairi} in \exref{bakverbs}.

@@

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
Further, while there are reflexes of \rc{a[p]} \qu{to be} in \ikpeng, only \rc{eti} forms are attested with first person inflectional prefixes \parencite[401]{gildea2018reconstructing}.

@@

\Cref{tab:ppekverbs} gives an overview of the first person forms of the seven verbs under discussion, along with our \PPek reconstruction.
The presence of the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø suggests to us that it may be cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}.
Indeed, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \cref{tab:pxinw}.
While it is by no means a regular sound change, it allows us to securely connect the two prefixes.
Similarly, the change of \rc{wɨ} to \bakairi \obj{u} is found in other correspondences, like \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
Thus, we reconstruct a \gl{1}\gl{s_a_} prefix \rc{w(ɨ)-} to \PPek, identical to \arara.

@@

For \qu{to go down}, we reconstruct \rc{ɨ} as the initial vowel rather than \rc{i}, see \cref{sec:godown}.
Further, the forms are not fully cognate; \textcite{meira2005southern} make no mention of a regular correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}.
However, the addition of a final \obj{ŋ} in \PXin is attested elsewhere: \begin{inlinelist}
\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} %\parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} %\parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} %\parencites[8]{gildea2007greenberg}[57]{alves2017arara}
\end{inlinelist} \parencites[8]{gildea2007greenberg}[56, 144, 57]{alves2017arara}[25, 270]{ikpengpacheco2001}.
Also, based on the fact \bakairi has generally lost much segmental material, we suggest that \bakairi \obj{gɨ} is another addition to a root \rc{ɨptə}, rather than a conservative form.

@@

The forms for \qu{to come} are not fully cognate, either: \ikpeng and \bakairi both show a reflex of the \PPek detransitivizer \rc{əd-} in combination with a root reconstructible as \rc{epɨ}.
In contrast, the \arara first person form is directly based on this root \rc{epɨ}.
However, other person values show a reflex of \rc{əd-} in \arara, too \exref{ara-123}.

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
On the other hand, both \ikpeng and \bakairi show reflexes of \rc{əd-ebɨ} throughout the whole paradigm.
Following the line of reasoning used by \textcite[114]{meira1998proto} (see also \cref{sec:taranoan}) for a similar pattern in the three Taranoan languages, we suggest that the idiosyncratic pattern in \arara is reconstructible to \PPek, and that \bakairi and \ikpeng independently regularized the paradigm to only use \rc{əd-ebɨ}.

The V-initial nature of \qu{to go} is evidenced in its Xinguan forms, while \bakairi \phon{\rc{wɨ}}{\obj{u}} resulted in a shift of the morpheme boundary, see also \cref{sec:go}.
The \ikpeng form \obj{aran} is made compatible with our reconstruction \rc{ɨtən} by considering that \ikpeng \obj{a} is an attested outcome of \rc{ə} (\PXin \rc{o}/\obj{e}), for example \begin{inlinelist}
 \item \obj{akari} \qu{dog} in \cref{tab:pxinw} above
 \item \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg}
 \item \obj{jaj} \qu{tree} \parencite[98]{ikpengpacheco2001} from \PC \rc{jəje}
 \end{inlinelist}.
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{\underline{ə}t\underline{ə}n}: \obj{\underline{e}r\underline{o}-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

Summing up, an innovative \gl{1}\gl{s_a_} marker \rc{k-} is reconstructible to \PPek.
Seven verbs can be reconstructed as having resisted this innovation, and preserving \gl{1}\gl{s_a_} \rc{w(ɨ)-} in \PPek.
In later, individual developments, \bakairi introduced \obj{k-} to \qu{to go down}, and \ikpeng to \qu{to go}.
Further, both languages regularized the paradigm of \qu{to come} to \rc{əd-epɨ} and introduced \obj{k-}.

\subsection{\PWai \rc{k-}}
\label{sec:waiwaian}
This extension was briefly discussed in \cref{sec:extensions_intro}; the new \gl{1}\gl{s_p_} prefix \rc{k-}, already innovated at the \PPar stage, was later extended to \gl{1}\gl{s_a_} in \PWai.
This created a unified \gl{1}\gl{s} category, reflected in both \hixka and \waiwai \pcref{tab:pwaireg}.

\input{floats/pwaireg}

%\ex<wais> Waiwaian \rc{eɸurka} \qu{to fall} (\gl{s_a_}) and \rc{wɨnɨkɨ} \qu{to sleep} (\gl{s_p_})\\
%\begin{tabular}[t]{@{}lllllll@{}}
%& \multicolumn{2}{l}{\PWai} & \multicolumn{2}{l}{\hixka}  & \multicolumn{2}{l}{\waiwai} \\
% & \gl{s_a_} & \gl{s_p_} & \gl{s_a_} & \gl{s_p_} & \multicolumn{2}{l}{\gl{s}} \\
%\gl{1} & \rc{\emp{k-}eɸurka-} & \rc{kɨ-wɨnɨkɨ-} & \obj{\emp{k-}ehurka-} & \obj{kɨ-nɨkɨ-} & \obj{\emp{k-}eɸɨrka-} & \obj{kɨ-wɨnk-} \\
%\gl{2} & \rc{m-eɸurka-} & \rc{o-wɨnɨkɨ-} & \obj{m-ehurka-} & \obj{o-wnɨkɨ-} & \obj{m-eɸɨrka-} & \obj{mɨ-wɨnk-} \\
%\gl{1+2} & \rc{t-eɸurka-} & \rc{tɨt-wɨnɨkɨ-} & \obj{t-ehurka-} & \obj{tɨ-nɨkɨ-} & \obj{t͡ʃ-eɸɨrka-} & \obj{tɨt-wɨnk-} \\
%\gl{3} & \rc{ɲ-eɸurka-} & \rc{nɨ-wɨnɨkɨ-} & \obj{ɲ-ehurka-} & \obj{nɨ-nɨkɨ-} & \obj{ɲ-eɸɨrka-} & \obj{nɨ-wɨnk-} \\
%\end{tabular}\\
%\parencites[510]{howard2001wrought}[189--191]{hixkaryanaderby1985}[209--211]{hawkins1953waiwai}[50]{waiwaihawkins1998}
%\xe
%
\waiwai \obj{ka} \qu{to say} does not take \obj{kɨ-}, but rather conservative \obj{wɨ-} \exref{hixɨ.wai-108}.
Its \hixka counterpart also takes a reflex of \gl{1}\gl{s_a_} \rc{w-}, which has the form \obj{ɨ-} \exref{hixɨ.hix-118}.
A formally identical prefix is found on transitive verbs in \gl{1}>\gl{3} scenarios in \hixka \exref{hixɨ.hix-17}, where it also corresponds to \waiwai \obj{w(ɨ)-} \exref{hixɨ.wai-159}.

\pex<hixɨ>
\a<wai-108> \waiwai \parencite[][71]{waiwaihawkins1998}\\
\begingl
\glpreamble wɨɨkekɲe//
\gla wɨ-ka-jakɲe//
\glb \gl{1}-say-\gl{pst}//
\glft \qu{I said.}//
\endgl
\a<hix-118> \hixka \parencite[][124]{hixkaryanaderby1985}\\
\begingl
\glpreamble roxehra nay hamɨ Kaywerye ɨkekonɨ//
\gla ro-ʃe-hɨra n-a-je hamɨ kajwerʲe ɨ-ka-jakonɨ//
\glb \gl{1}-\gl{des}-\gl{neg} \gl{3}-be-\gl{npst}.\gl{uncert} \gl{evid} K. \gl{1}\gl{s_a_}-say-\gl{rem}.\gl{cont}//
\glft \qu{I said (to myself), “Kaywerye evidently doesn't like me”.}//
\endgl
\a<hix-17> \hixka \parencite[][191]{hixkaryanaderby1985}\\
\begingl
\gla ɨ-koroka-no//
\glb \gl{1}>\gl{3}-wash-\gl{imm}//
\glft \qu{I washed him.}//
\endgl
\a<wai-159> \waiwai \parencite[][192]{waiwaihawkins1998}\\
\begingl
\glpreamble wîyesî//
\gla wɨ-jo-jasɨ//
\glb \gl{1}>\gl{3}-boil-\gl{npst}//
\glft \qu{I will boil it.}//
\endgl
\xe
%
This correspondence allows us to establish \hixka \obj{ɨ-} a reflex of \rc{wɨ-}, similar to \ikpeng \pcref{sec:pekodian}.
Notably, \textcite{hixkaryanaderby1985} does not see this \obj{ɨ-} as an irregular \gl{1}\gl{s_a_}, but as the regular \gl{1}>\gl{3} marker, because he considers \hixka \obj{ka} \qu{to say} to be transitive, an issue which will be discussed in \cref{sec:say}.

There are three more verbs which did not take innovative \rc{k-} in \PWai, shown alongside \rc{ka} \qu{to say} in \cref{tab:pwaiverbs}.
While the two roots for \qu{to be} are straightforwardly reconstructible, \qu{to go} is somewhat of a special case.
While \hixka has the expected \obj{ɨ-}, \waiwai seems to have combined innovative \gl{1}\gl{s_a_} \obj{k-} with the old \rc{w-}, an analysis also considered by \textcite[90]{gildea1998}.
Alternatively, this form was influenced by deverbalized forms of \qu{to go}, where a reflex of the \gl{s_a_} class marker \rc{w-} has become fossilized \exref{waiwto}.

@@

\subsection{\PTir \rc{t-}}
\label{sec:taranoan}
The Taranoan languages were already grouped together by \textcite{girard1971proto}, consisting of closely related \trio and \akuriyo (subsumed here under the moniker Tiriyoan), and the more distant member \carijo.
\textcite{meira1998proto} provides an extensive phonological, morphological, and lexical reconstruction of \PTar.
The \setone paradigms of \trio and \akuriyo contain an interesting puzzle: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places.

>>>>

\section{The innovative \gl{1}\gl{s_a_} markers}
\label{sec:extensions}
As stated in \cref{sec:extensions_intro}, the person marker extensions which did not spread through the entire lexicon all have in common that they feature innovative first person markers on verbs that are (at least historically) members of the \gl{s_a_} class.
There are six attested such innovations, three of which can be reconstructed to intermediate proto-languages.
The new \gl{1}\gl{s_a_} prefix is formally identical to the \gl{1+2} marker (\PC \rc{k-}) in three cases, to the \gl{1}\gl{s_p_} marker (\PC \rc{u(j)-}) in two cases, and to the \gl{1}>\gl{3} marker (\PC \rc{t-}) in one case.
\cref{sec:pekodian} investigates the innovation of \rc{k-} in \PPek, reflected in the three daughter languages \arara, \ikpeng, and \bakairi.
\cref{sec:waiwaian} takes a closer look at the extension of \rc{k-} in \PWai, which was briefly shown in \cref{sec:extensions_intro}.
\cref{sec:taranoan} concerns the extension of \rc{t-} in \PTir (or \PTar), reflected in modern \trio and \akuriyo.
\crefrange{sec:akuriyo}{sec:yukpa} look at innovative first person markers which are only attested in single modern languages:
\obj{k-} in \akuriyo, and \obj{j-} in \carijo and \yukpa.

\subsection{\PPek \rc{k-}}
\label{sec:pekodian}
The Pekodian branch was suggested by \textcite{meira2005southern}, as the result of fieldwork on \bakairi by Meira and the availability of more material on \ikpeng. % \parencites{pacheco1997relativizacao}{ikpengpacheco1997}{campetela1997analise}{pacheco1998nominalizacao}{ikpengpacheco2001}{campetela2002aspectos}{pacheco2002aspectos}{pacheco2003intransitivos}.
It consists of closely related \arara and \ikpeng, with \bakairi as a more distant member.
\textcite{meira2005southern} focused on phonological and lexical properties, so no reconstructive work on \PPek morphosyntax can be found in the literature.
All three Pekodian languages have a regular \gl{1}\gl{s_a_} marker \obj{k-}, as evidenced by the paradigms in \cref{tab:pekreg}.\footnote{In \cref{tab:pekreg} and the remainder of this paper, we omit any possible TAM suffixes, since our focus lies on the prefixes and stems, and since full paradigms containing the same TAM suffix are rarely found.}
We can thus reconstruct a \PPek \gl{1}\gl{s_a_} marker \rc{k-}.

@@

\begin{multicols}{2}
\ex<arairr> \arara \parencite[153]{alves2017arara} \\
\begin{tabular}[t]{@{}ll@{}}
\obj{wɨ-genɨ} & \qu{I said}\\
\obj{w-it͡ʃinɨ} & \qu{I was, lied down}\\
\obj{w-ebɨnɨ} & \qu{I came}\\
\obj{w-ibɨnɨ} & \qu{I bathed}\\
\obj{w-iptoŋrɨ} & \qu{I went down}\\
\obj{w-ɨdolɨ} & \qu{I went}\\
\end{tabular}
\xe
%
\ex<araap> \arara \parencite[200]{alves2017arara}\\
\begin{tabular}[t]{@{}ll@{}}
\gl{1} & \obj{w-aptam} \qu{when/if I was}\\
\gl{2} & \obj{m-od-aptam}\\
\gl{1+2} & \obj{kud-aptam}\\
\gl{3} & ∅\obj{-aptam}\\
\end{tabular}
\xe
\end{multicols}

In his brief discussion of \bakairi verbal person marking, \textcite{meira2003bakairi} reports the existence of two subclasses of \gl{s_a_} verbs, one taking first person \obj{w-}, and one \obj{k-}.
The verb used to illustrate the first group is \obj{i} \qu{to bathe} \exref{bak-33}, contrasting with regular \obj{əku} \qu{to go up} in \cref{tab:pekreg} above.

\ex<bak-33>\bakairi \parencite[][4]{meira2003bakairi}\\
\begingl
\gla w-i-də//
\glb \gl{1}\gl{s_a_}-bathe-\gl{imm}//
\glft \qu{I bathed}//
\endgl
\xe
%
Of course, \qu{to bathe} is also found in the \obj{w-}list for \arara.
While \textcite[4]{meira2003bakairi} does list \obj{ge} \qu{to say}, \obj{tə} \qu{to go}, and \obj{əe(wɨ)} \qu{to come} as examples of \gl{s_a_} verbs, he does not indicate whether they belong to the class of \gl{s_a_}-1 verbs, with first person \obj{k-}, or the \gl{s_a_}-2 verbs, with \obj{w-}.%
\footnote{
It should be noted at this point that \textcite{meira2003bakairi} indicates that the same verbs which take first person \obj{w-} in \bakairi also take a \gl{1+2} marker \obj{k-}.
However, this marker is only illustrated for \qu{to bathe}, both by \textcite{meira2003bakairi} and \textcite{von1892bakairi}.
Given the lack of data for other verbs, we will not further discuss this potential additional pattern.
If the characterization by \citeauthor{meira2003bakairi} is accurate, then the pattern is fully parallel to the distribution of the first person prefixes.
}
Luckily, while \textcite{von1892bakairi} did not accurately record all phonemic distinctions in \bakairi \parencite{meira2005bakairi}, he does provide inflected forms of cognates to the \arara verbs in \exref{arairr}.
We present them in \exref{bakverbs} according to our current understanding of \bakairi phonology and verbal morphology, based on \textcites{wheatley1969bakairi}{meira2003bakairi}{meira2005bakairi}{franchetto2016classes}.

@@

All available descriptions of the third Pekodian language, \ikpeng, list \obj{k-} as the only \gl{1}\gl{s_a_} marker \parencites[55]{ikpengpacheco1997}[105]{campetela1997analise}[64]{ikpengpacheco2001}[205]{alves2013verbo}.
However, most \ikpeng cognates of the \arara verbs with \gl{1}\gl{s_a_} \obj{w-} actually do not take \obj{k-}, but rather \obj{ɨ-} or Ø, as shown in \exref{ikpw}.
The exception is \qu{to go}, which has \obj{k-} \exref{ikp-170}.
There is a formally identical \ikpeng cognate of \arara \obj{iptoŋ} \qu{to go down}, but no first person forms are attested \perscommpar{Angela Chagas}.
Further, while there are reflexes of \rc{a[p]} \qu{to be} in \ikpeng, only \rc{eti} reflexes seem to occur with first person inflectional prefixes \parencite[401]{gildea2018reconstructing}.

@@

\Cref{tab:ppekverbs} gives an overview of the first person forms of the seven verbs under discussion, along with our \PPek reconstruction.
The presence and distribution of the \ikpeng \gl{1}\gl{s_a_} marker \obj{ɨ-}/Ø suggests that it is cognate with \arara \gl{1}\gl{s_a_} \obj{w(ɨ)-}.
Indeed, \PXin \rc{w} is attested as sometimes being lost in \ikpeng, as evidenced by the correspondences in \cref{tab:pxinw}.
While it is by no means a regular sound change, it allows us to securely connect the two prefixes.
Similarly, the supposed change of \rc{wɨ} to \bakairi \obj{u} is found in other correspondences, like \obj{udo} \parencite{meira2005southern} from \PC \rc{wɨtoto} \qu{person} \parencite[4]{gildea2007greenberg}.
Thus, we reconstruct a \gl{1}\gl{s_a_} prefix \rc{w(ɨ)-} to \PPek, identical to the \arara one in form and distribution.

@@

For \qu{to go down}, we reconstruct \rc{ɨ} as the initial vowel rather than \rc{i}, see \cref{sec:godown}.
Further, the forms are not fully cognate; \textcite{meira2005southern} make no mention of a regular correspondence between \bakairi \obj{gɨ} and \ikpeng \obj{ŋ}.
However, the addition of a final \obj{ŋ} in \PXin is attested elsewhere,%
\footnote{
\begin{inlinelist}
\item \PC \rc{əne} \qu{to see}, \arara and \ikpeng \obj{eneŋ} %\parencites[8]{gildea2007greenberg}[56]{alves2017arara}[25]{ikpengpacheco2001}
\item \PC \rc{əta} \qu{to hear}, \arara \obj{taŋ}, \ikpeng \obj{iraŋ} %\parencites[8]{gildea2007greenberg}[144]{alves2017arara}[270]{ikpengpacheco2001}
\item \PC \rc{ənə} \qu{to eat meat}, \arara \obj{oŋoŋ} \qu{to bite} %\parencites[8]{gildea2007greenberg}[57]{alves2017arara}
\end{inlinelist} \parencites[8]{gildea2007greenberg}[56, 144, 57]{alves2017arara}[25, 270]{ikpengpacheco2001}.}
and based on the fact \bakairi has generally lost much segmental material, we suggest that both \obj{ŋ} and \obj{gɨ} are later additions to a root \rc{ɨptə}.

@@

The forms for \qu{to come} are not fully cognate, either: \ikpeng and \bakairi both show a reflex of the \PPek detransitivizer \rc{əd-} in combination with a root reconstructible as \rc{epɨ}.
In contrast, the \arara first person form is directly based on this root \rc{epɨ}.
However, reflexes of \rc{əd-epɨ} can be found elsewhere in the \arara paradigm \exref{ara-123}.

\ex<ara-123>\arara \parencite[][150]{alves2017arara}\\
\begingl
\gla m-odebɨ-nɨ//
\glb \gl{2}\gl{s_a_}-come-\gl{rec}//
\glft \qu{You came.}//
\endgl
\xe
%
On the other hand, both \ikpeng and \bakairi show reflexes of \rc{əd-ebɨ} throughout the whole paradigm.
Following the line of reasoning used by \textcite[114]{meira1998proto} (see also \cref{sec:taranoan}) for a similar pattern in the three Taranoan languages, we suggest that the idiosyncratic pattern in \arara is reconstructible to \PPek, and that \bakairi and \ikpeng independently regularized the paradigm to only use \rc{əd-epɨ}.

The V-initial nature of \qu{to go} is evidenced in its Xinguan forms, while \bakairi \phon{\rc{wɨ}}{\obj{u}} resulted in a shift of the morpheme boundary (see also \cref{sec:go}).
The \ikpeng form \obj{aran} is compatible with our reconstruction \rc{ɨtən}, when one considers that \ikpeng \obj{a} is an attested outcome of \rc{ə}.\footnote{\begin{inlinelist}
 \item \obj{akari} \qu{dog} in \cref{tab:pxinw} above
 \item \obj{anma} \qu{path} \parencite[24]{ikpengpacheco2001} from \PC \rc{ətema} \parencite[12]{gildea2007greenberg}
 \item \obj{jaj} \qu{tree} \parencite[98]{ikpengpacheco2001} from \PC \rc{jəje}
 \end{inlinelist}.}
This attested change of \rc{ə} to \obj{a} need only be preceded by a assimilatory lowering of initial \rc{ɨ} to \rc{ə}, to yield the form \obj{aran} from \rc{ɨtən}.
Other \ikpeng reflexes of \qu{to go} offer evidence for the suggested intermediate stage \rc{\underline{ə}t\underline{ə}n}: \obj{\underline{e}r\underline{o}-lɨ} \qu{s/he went} \parencite[25]{ikpengpacheco2001}.

Summing up, an innovative \gl{1}\gl{s_a_} marker \rc{k-} is reconstructible to \PPek.
Seven verbs can be reconstructed as having resisted this innovation, and preserving \gl{1}\gl{s_a_} \rc{w(ɨ)-} in \PPek.
In later, individual developments, \bakairi extended \obj{k-} to \qu{to go down}, and \ikpeng to \qu{to go}.
Further, both languages regularized the paradigm of \qu{to come} to \rc{əd-epɨ}, accompanied by the extension of \obj{k-}.

\subsection{\PWai \rc{k-}}
\label{sec:waiwaian}
This extension was briefly discussed in \cref{sec:extensions_intro}; the new \gl{1}\gl{s_p_} prefix \rc{k-}, already innovated at the \PPar stage, was later extended to \gl{1}\gl{s_a_} in \PWai.
This created a unified \gl{1}\gl{s} category, reflected in both \hixka and \waiwai \pcref{tab:pwaireg}.
%
\input{floats/pwaireg}
%
%\ex<wais> Waiwaian \rc{eɸurka} \qu{to fall} (\gl{s_a_}) and \rc{wɨnɨkɨ} \qu{to sleep} (\gl{s_p_})\\
%\begin{tabular}[t]{@{}lllllll@{}}
%& \multicolumn{2}{l}{\PWai} & \multicolumn{2}{l}{\hixka}  & \multicolumn{2}{l}{\waiwai} \\
% & \gl{s_a_} & \gl{s_p_} & \gl{s_a_} & \gl{s_p_} & \multicolumn{2}{l}{\gl{s}} \\
%\gl{1} & \rc{\emp{k-}eɸurka-} & \rc{kɨ-wɨnɨkɨ-} & \obj{\emp{k-}ehurka-} & \obj{kɨ-nɨkɨ-} & \obj{\emp{k-}eɸɨrka-} & \obj{kɨ-wɨnk-} \\
%\gl{2} & \rc{m-eɸurka-} & \rc{o-wɨnɨkɨ-} & \obj{m-ehurka-} & \obj{o-wnɨkɨ-} & \obj{m-eɸɨrka-} & \obj{mɨ-wɨnk-} \\
%\gl{1+2} & \rc{t-eɸurka-} & \rc{tɨt-wɨnɨkɨ-} & \obj{t-ehurka-} & \obj{tɨ-nɨkɨ-} & \obj{t͡ʃ-eɸɨrka-} & \obj{tɨt-wɨnk-} \\
%\gl{3} & \rc{ɲ-eɸurka-} & \rc{nɨ-wɨnɨkɨ-} & \obj{ɲ-ehurka-} & \obj{nɨ-nɨkɨ-} & \obj{ɲ-eɸɨrka-} & \obj{nɨ-wɨnk-} \\
%\end{tabular}\\
%\parencites[510]{howard2001wrought}[189--191]{hixkaryanaderby1985}[209--211]{hawkins1953waiwai}[50]{waiwaihawkins1998}
%\xe
%
Not all verbs were affected by this extension; \waiwai \obj{ka} \qu{to say} does not take \obj{kɨ-}, but rather conservative \obj{wɨ-} \exref{hixɨ.wai-108}.
Its \hixka counterpart has a prefix \obj{ɨ-} \exref{hixɨ.hix-118}, a potential reflex of \gl{1}\gl{s_a_} \rc{w(ɨ)-}.
A formally identical prefix occurs in \gl{1}>\gl{3} scenarios in \hixka \exref{hixɨ.hix-17}, which regularly corresponds to \waiwai \obj{w(ɨ)-} \exref{hixɨ.wai-159}.

\pex<hixɨ>
\a<wai-108> \waiwai \parencite[][71]{waiwaihawkins1998}\\
\begingl
\glpreamble wɨɨkekɲe//
\gla wɨ-ka-jakɲe//
\glb \gl{1}-say-\gl{pst}//
\glft \qu{I said.}//
\endgl
\a<hix-118> \hixka \parencite[][124]{hixkaryanaderby1985}\\
\begingl
\glpreamble roxehra nay hamɨ Kaywerye ɨkekonɨ//
\gla ro-ʃe-hɨra n-a-je hamɨ kajwerʲe ɨ-ka-jakonɨ//
\glb \gl{1}-\gl{des}-\gl{neg} \gl{3}-be-\gl{npst}.\gl{uncert} \gl{evid} K. \gl{1}\gl{s_a_}-say-\gl{rem}.\gl{cont}//
\glft \qu{I said (to myself), “Kaywerye evidently doesn't like me”.}//
\endgl
\a<hix-17> \hixka \parencite[][191]{hixkaryanaderby1985}\\
\begingl
\gla ɨ-koroka-no//
\glb \gl{1}>\gl{3}-wash-\gl{imm}//
\glft \qu{I washed him.}//
\endgl
\a<wai-159> \waiwai \parencite[][192]{waiwaihawkins1998}\\
\begingl
\glpreamble wîyesî//
\gla wɨ-jo-jasɨ//
\glb \gl{1}>\gl{3}-boil-\gl{npst}//
\glft \qu{I will boil it.}//
\endgl
\xe
%
This correspondence allows us to establish \hixka \obj{ɨ-} a reflex of \rc{w(ɨ)-}, similar to \ikpeng \pcref{sec:pekodian}.
Notably, \textcite{hixkaryanaderby1985} does not see this \obj{ɨ-} as an irregular \gl{1}\gl{s_a_}, but as the regular \gl{1}>\gl{3} marker, because he considers \hixka \obj{ka} \qu{to say} to be transitive, an issue which will be discussed in \cref{sec:say}.

There are three more verbs which did not take innovative \rc{k-} in \PWai, shown alongside \rc{ka} \qu{to say} in \cref{tab:pwaiverbs}.
The two roots for \qu{to be} are straightforwardly reconstructible, whereas \qu{to go} is somewhat of a special case.
While \hixka has the expected \obj{ɨ-}, \waiwai seems to have combined innovative \obj{k-} with the old \rc{w-}, an analysis also considered by \textcite[90]{gildea1998}.
Alternatively, this form was influenced by deverbalized forms of \qu{to go}, where a reflex of the \gl{s_a_} class marker \rc{w-} has become fossilized \exref{waiwto}.

@@

\subsection{\PTir \rc{t-}}
\label{sec:taranoan}
The Taranoan languages were already grouped together by \textcite{girard1971proto}, consisting of closely related \trio and \akuriyo (subsumed here under the moniker Tiriyoan), and the more distant member \carijo.
\textcite{meira1998proto} provides an extensive phonological, morphological, and lexical reconstruction of \PTar.
The \setone paradigms of \trio and \akuriyo contain an interesting puzzle: \PC \gl{1}>\gl{3} \rc{t-} and \gl{1}\gl{s_a_} \rc{w-} seem to have switched places.
This resulted in a regular \gl{1}\gl{s_a_} marker \envr{\rc{t͡ʃ-}}{\obj{e}}, \envr{\rc{t-}}{\obj{ə}} \pcref{tab:ptirreg}; the latter allomorph was subsequently replaced by \obj{k-} in \akuriyo \pcref{sec:akuriyo}.
